The late 1800's debate over the age of the Earth: an early attempt of interdisciplinary science

### 2023-08-16 ms.tex
Sec 3

\title{Lord Kelvin's debate over the age of the Earth: an early attempt of interdisciplinary science}
\author{Nicholas Wogan}
\maketitle

\begin{abstract}
  In the 1860s, Lord Kelvin used his understanding of physics and thermodynamics to estimate the age of the Earth. He brought these calculations to the attention of geologists, arguing against the uniformitarian view of an indefinitely old Earth and troubling Darwin, who wanted vast periods of history for his theory of natural selection. The ensuing debate, involving Lyell's geology, Darwin's biology, and Kelvin's physics, is an early attempt of interdisciplinary science. However, Kelvin's quantitative arguments for a young Earth were not entertained or understood by large portion of the geologic community for about a decade. Here, I argue that one reason this interdisciplinary science failed to proceed smoothly is because, at the time, there were big cultural divisions between the quantitative physical sciences and the more qualitative geologic and biological sciences. Today, some of the big unsolved problems in science will likely require collaborations between many different disciplines. To solve these big problems, perhaps we can learn from the failed 1860s debate over the age of the Earth, by striving to break down cultural and tribal barriers between different fields of study.
\end{abstract}

@@

\section{Lessons from history} \label{sec:3}

@@

>>>>

\title{The late 1800's debate over the age of the Earth: an early attempt of interdisciplinary science}
\author{Nicholas Wogan}
\maketitle

@@

\section{A reason why late 1800s interdisciplinary science was unsuccessful and implications for modern research} \label{sec:3}

The debate over the age of the Earth was one of the earliest attempts of interdisciplinary science. The problem involved Kelvin's thermodynamic arguments, the geologic record, as well as Darwin's theory of biological evolution. However, this interdisciplinary experiment was not very successful. For years after his initial 1862 arguments for a young Earth, Kelvin was unable to convince geologists to take his calculations seriously. Also, Kelvin seemed to value his thermodynamics far more than evidence preserved in the stratigraphic record. Why couldn't any physicist, geologist or biologist consider all the evidence (geologic, or physics-based) to estimate Earth's age?

There are likely many reasons why interdisciplinary science failed. I believe one important reason was that, at the time, different disciplines were culturally tied to specific kinds of scientific evidence. Thanks to Lyell's paradigm of uniformitarianism, the late 1800s geologist took a qualitative approach, studying the Earth by solely examining piles of strata. This method for understanding the world was part of what it meant to be a geologist. In contrast, a physicist was defined by a mathematical or quantitative approach. Given this cultural division between the disciplines, no researcher was in a position to consider all evidence on equal footing.

@@

In the 1860s, Lord Kelvin used his understanding of physics and thermodynamics to estimate the age of the Earth. He brought these calculations to the attention of geologists, arguing against the uniformitarian view of an indefinitely old Earth and troubling Darwin, who wanted vast periods of history for his theory of natural selection. The ensuing debate, involving Lyell's geology, Darwin's biology, and Kelvin's physics, is an early attempt of interdisciplinary science. However, Kelvin's quantitative arguments for a young Earth were not entertained or understood by large portion of the geologic community for about a decade. Here, I argue that one reason this interdisciplinary science failed to proceed smoothly is because, at the time, there were big cultural divisions between the quantitative physical sciences and the more qualitative geologic and biological sciences. Today, some of the big unsolved problems in science will likely require collaborations between many different disciplines. To solve these big problems, perhaps we can learn from the failed 1860s debate over the age of the Earth, by striving to break down cultural and tribal barriers between different fields of study.
